---
title: "Hierarchical Linear Models"
author: "John D. Lee and Linda Ng Boyle"
date: 10/1/2018
output:
  ioslides_presentation:
    widescreen: yes
  beamer_presentation: default
---
<style>
pre {
  font-size: 18px;
  line-height: 1.05;
}
</style>
  
   
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache = TRUE)

#### Use caret to develop, tune and validate a range of data mining models ####
# http://www.jstatsoft.org/v28/i05/paper

library(tidyverse)
library(skimr)
library(lme4)
library(lmerTest)
library(psycho)
set.seed(1)
rm(list = ls(all = TRUE))
```


## A Common Challenge|Repeated measures 
-- Multiple measures from the same person who might also be part of a group

-- Multiple exposures across people to the same stimuli

-- Repeated measures ANOVA accommodates correlated observations

-- Multi-level general linear models go further and explicitly model grouping effects 

-- Addresses Simpson's paradox, where individual trends don't reflect the aggregate trend


## Simpson's Paradox
```{r simpons_pardox, echo = TRUE, warnings=FALSE, fig.height=5.25, fig.width=6.5, echo=FALSE}
library(scales)
set.seed(123)
n = 1000

Education = rbinom(n, 2, 0.5)
Neuroticism = rnorm(n) + Education
Salary = Education * 2 + rnorm(n) - Neuroticism * 0.3
Salary = sample(10000:11000,1) + rescale(Salary, to = c(0, 100000))
Neuroticism = rescale(Neuroticism, to = c(0, 7))
Education = factor(Education, labels = c("Low", "Medium", "High"))

data <- data.frame(
  Salary,
  Neuroticism,
  Education
)

ggplot(data, aes(Neuroticism, Salary)) +
  #geom_point(alpha = .5, size = 4) + 
  geom_point(aes(col = Education), alpha = .4, size = 2) + 
  geom_smooth(aes(col = Education), method = 'lm') +
  geom_smooth(method = 'lm') +
  theme(legend.background = element_rect(fill = "transparent"),
        legend.justification = c(0, 1),
        legend.position = c(0, 1))+
  theme_grey(base_size = 14)
```
<span style="font-size:.5em">http://rpubs.com/lakenp/simpsonsparadox</span>


## Useful packages
-- ``library(tidyverse)`` for ggplot2, dplyr

-- ``library(skimr)`` for skim 

-- ``library(lme4)`` for mixed effect, multi-level general linear models

-- ``library(lmerTest)`` for extracting p-values

-- ``library(psycho)`` for extracting APA reportable sentences


## Effect of Sleep Disruption
```{r facet_scatterplot_lm, echo = TRUE, warnings=FALSE, fig.height=3.75, fig.width=10.25}
sleep.df = sleepstudy

ggplot(sleep.df, aes(Days, Reaction)) + 
   geom_point() +
   geom_smooth(method = "lm") +
   facet_grid(.~Subject)
```


## Uniform Effect of Sleep Disruption?  
```{r facet_lm, echo = TRUE, warnings=FALSE, fig.height=4}
sleep.fit = lm(Reaction ~ Days, sleep.df)

sleep.df$prediction = predict(sleep.fit)

summary(sleep.fit)
```


## Uniform Effect of Sleep Disruption? 
```{r facet_scatterplot_overalllm, echo = TRUE, warnings=FALSE, fig.height=4, fig.width=10.25}

ggplot(sleep.df, aes(Days, Reaction)) + 
  geom_point(alpha = .4, size = 1) +
  geom_point(aes(y=prediction), shape = 21) +
  geom_line(aes(y=prediction)) + 
  facet_grid(.~Subject)
```




## Random Effect for Intercept
```{r lmer_int, echo = TRUE, warnings=FALSE, fig.height=4, fig.width=10.25}
sleep_int.fit = lmer(Reaction ~ Days + (1|Subject), sleep.df)
summary(sleep_int.fit)

sleep.df$prediction_int = predict(sleep_int.fit)

```


## Random Effect for Intercept
```{r facet_scatterplot_lm_int, echo = TRUE, warnings=FALSE, fig.height=4, fig.width=10.25}

ggplot(sleep.df, aes(Days, Reaction)) + 
  geom_point(alpha = .4, size = 1) +
  geom_point(aes(y=prediction_int), shape = 21) +
  geom_line(aes(y=prediction_int, group = Subject)) + 
  facet_grid(.~Subject)

```

## Random Effect for Intercept and Slope
```{r lmer_int_slope, echo = TRUE, warnings=FALSE, fig.height=4, fig.width=10.25}

sleep_int_slope.fit = lmer(Reaction ~ Days + (Days|Subject), sleep.df)
summary(sleep_int_slope.fit)

sleep.df$prediction_int_slope = predict(sleep_int_slope.fit)

```


## Random Effects for Intercept and Slope
```{r facet_scatterplot_lm_int_slope, echo = TRUE, warnings=FALSE, fig.height=4, fig.width=10.25}

ggplot(sleep.df, aes(Days, Reaction)) + 
  geom_point(alpha = .4, size = 1) +
  geom_line(aes(y=prediction_int_slope, group = Subject)) + 
  geom_point(aes(y=prediction_int_slope), shape = 21) +
  facet_grid(.~Subject)

```


## Summary for Publication
-- The "psycho" package provides functions for summarizing model output

-- It creates a publication-ready summary table

-- It can even create sentences describing the output

-- Parameter estimates can be extracted for plotting

```{r publication_summary1, echo = TRUE, warning=FALSE, fig.height=3,out.width=11}
library(psycho)
analyze(sleep_int_slope.fit, CI = 95) %>% 
  summary(results, round = 2) %>% 
  mutate(p = format_p(p)) %>% 
  kable()


```


## Summary for Publication|psycho can create sentences describing the output
```{r publication_summary2, echo = TRUE, warning=FALSE}
analyze(sleep_int_slope.fit, CI = 95)
```

-The overall model predicting Reaction (formula = Reaction ~ Days + (Days | Subject)) has an total explanatory power (conditional R2) of 79.92%, in which the fixed effects explain 27.87% of the variance (marginal R2). 

-The model's intercept is at 251.41 (SE = 6.82, 95% CI [237.68, 265.13]). Within this model:

-The effect of Days is significant (beta = 10.47, SE = 1.55, 95% CI [7.36, 13.58], t(17) = 6.77, p < .001) and can be considered as medium (std. beta = 0.54, std. SE = 0.079).


## Summary for Publication|Parameter estimates can also be extracted for plotting

```{r publication_summary3, echo = TRUE, warning=FALSE, fig.height=3.}

CI.lme.df = analyze(sleep_int_slope.fit)$summary

ggplot(CI.lme.df, aes(Variable)) +
    geom_hline(yintercept = 0, alpha = .5)+
    geom_pointrange(aes(y = Coef, ymin= CI_lower, ymax = CI_higher)) +
    coord_flip() +
    labs(x = "", y = expression(Coefficient~and~95^{th}~CI))
```


## Exercise|Multi-level model fitting

-- Fit a multi-level model with Subject as a random variable

-- Use the ``analyze`` command from psych to extract an summmary of effects

-- Plot the effect estimates

```{r glmfit, echo = FALSE, warning=FALSE, fig.height=3.}

sleep_int.fit = lmer(Reaction ~ Days + (Days|Subject), sleep.df)

CI.lme.df = analyze(sleep_int.fit)$summary

ggplot(CI.lme.df, aes(Variable)) +
    geom_hline(yintercept = 0, alpha = .5)+
    geom_pointrange(aes(y = Coef, ymin= CI_lower, ymax = CI_higher)) +
    coord_flip() +
    labs(x = "", y = expression(Coefficient~and~95^{th}~CI))

```


## A Common Challenge|Repeated measures 
-- Multiple measures from the same person who might also be part of a group

-- Multiple exposures across people to the same stimuli

-- Repeated measures ANOVA accommodates correlated observations

-- Multi-level general linear models go further and explicitly model grouping effects

-- Addresses Simpson's paradox, where individual trends don't reflect the aggregate trend


